Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-06-18 paper/paper.tex
changed Bayesian -> probabilistic

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies. The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods. However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain. In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.  
Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition. Next, we demonstrate how the nodes of the fitted GTM can be interpreted as unique spectral endmembers. Using extracted endmembers together with the normalized spectral similarity score, we are able to efficiently map the abundance of near shore algae as well as the evolution of a rhodamine tracer dye used to simulate water contamination by a localized source. 
%This study highlights the value of combining UAV-based hyperspectral imaging with unsupervised methods for...
}

@@

% neurons that wire together, fire together
The self-organizing map (SOM) developed by Teuvo Kohonen is an unsupervised machine learning method which non-linearly maps high dimensional data to the nodes of a two-dimensional grid \cite{kohonen-som-1}. By preserving the topological relationship between nodes during training, the SOM ensures that similar spectral signatures are mapped together such that related HSI pixels naturally cluster together. This presents an attractive compromise by enabling the simultaneous visualization of HSI data and endmember extraction via the weight vector associated with each SOM node \cite{cantero2004analysis, duran2007time,som-hsi}. When reference data are available, the SOM can be utilized to provide semi-supervised labeling of HSI spectra \cite{riese2019supervised}. Furthermore, Danielsen et al. demonstrated that the dimensionality reduction offered by the SOM can even be used for on-board data compression of HSIs acquired by a CubeSat \cite{danielsen2021self}. Despite these clear capabilities, the SOM relies on a heuristic training algorithm with hyperparameters that can be challenging to tune and offers no direct probabilistic interpretation. To address these limitations, Bishop et al. developed the generative topographic mapping (GTM), a Bayesian latent-variable model inspired by the SOM \cite{gtm-bishop-1}. The GTM has been utilized in a variety of domains including drug design and chemical data visualization but has yet to see adoption for the analysis of hyperspectral imagery \cite{kireeva2012generative, gaspar2015chemical, horvath2019generative}.

@@

Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$ as illustrated in Fig~\ref{fig:gtm-diagram}. However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin. To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$. For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many i.i.d. records, this choice yields a log-likelihood function of the form
\begin{equation}\label{eq:llh}
    \mathcal{L}(W, \beta) = \sum_n^N \ln \left(\dfrac{1}{K}\sum_k^K p(\mathbf{x}_n \mid \xi_k, W, \beta) \right)
\end{equation}
which can be maximized to yield optimal $W$ and $\beta^{-1}$.

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = 2P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space. For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the embedding space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\authorcontributions{Methodology, J.W. and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W. and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L. All authors have read and agreed to the published version of the manuscript.
}

@@

\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required. Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}

\acknowledgments{Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.  Christopher Simmons is gratefully acknowledged for his computational support. We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite. Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage. Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team. The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the %Please ensure meaning has been retained. 
	NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies. The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods. However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain. In this study, we demonstrate how the Generative Topographic Mapping (GTM), a probabilistic realization of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.  Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition. Next, we demonstrate how the nodes of the fitted GTM can be interpreted as unique spectral endmembers. Using extracted endmembers together with the normalized spectral similarity score, we are able to efficiently map the abundance of near shore algae as well as the evolution of a rhodamine tracer dye used to simulate water contamination by a localized source. 
%This study highlights the value of combining UAV-based hyperspectral imaging with unsupervised methods for...
}

@@

% neurons that wire together, fire together
The self-organizing map (SOM) developed by Teuvo Kohonen is an unsupervised machine learning method which non-linearly maps high dimensional data to the nodes of a two-dimensional grid \cite{kohonen-som-1}. By preserving the topological relationship between nodes during training, the SOM ensures that similar spectral signatures are mapped together such that related HSI pixels naturally cluster together. This presents an attractive compromise by enabling the simultaneous visualization of HSI data and endmember extraction via the weight vector associated with each SOM node \cite{cantero2004analysis, duran2007time,som-hsi}. When reference data are available, the SOM can be utilized to provide semi-supervised labeling of HSI spectra \cite{riese2019supervised}. Furthermore, Danielsen et al. demonstrated that the dimensionality reduction offered by the SOM can even be used for on-board data compression of HSIs acquired by a CubeSat \cite{danielsen2021self}. Despite these clear capabilities, the SOM relies on a heuristic training algorithm with hyperparameters that can be challenging to tune and offers no direct probabilistic interpretation. To address these limitations, Bishop et al. developed the generative topographic mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-bishop-1}. The GTM has been utilized in a variety of domains including drug design and chemical data visualization but has yet to see adoption for the analysis of hyperspectral imagery \cite{kireeva2012generative, gaspar2015chemical, horvath2019generative}.

@@

Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$ as illustrated in Fig~\ref{fig:gtm-diagram}. However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin. To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$. For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log-likelihood function of the form
\begin{equation}\label{eq:llh}
    \mathcal{L}(W, \beta) = \sum_n^N \ln \left(\dfrac{1}{K}\sum_k^K p(\mathbf{x}_n \mid \xi_k, W, \beta) \right)
\end{equation}
which can be maximized to yield optimal $W$ and $\beta^{-1}$.

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the latent space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted. For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the latent space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\authorcontributions{Methodology, J.W. and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., G.B., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W. and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L. All authors have read and agreed to the published version of the manuscript.
}

@@

\dataavailability{The original data presented in the study are made openly available via the Open Storage Network at \url{https://ncsa.osn.xsede.org/ees230012-bucket01/RobotTeam/unsupervised/gtm-data/}. The open-source implementation of the GTM used in this study is available at \url{https://github.com/john-waczak/GenerativeTopographicMapping.jl}.}

\acknowledgments{Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.  Christopher Simmons is gratefully acknowledged for his computational support. We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite. Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage. Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team. The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}
